Learning and Evaluation of \strips\ Action Models with Classical Planning

### 2018-05-02 doc/aij18/compilation.tex
new updates to the AIJ18 paper

Our approach for addressing a $\Lambda$ learning task is compiling it into a classical planning task $P_{\Lambda}$ with conditional effects. A planning compilation is a suitable approach because computing a solution for $\Lambda$ involves, not only determining the \strips\ action model $\mathcal{M}'$, but also the {\em unobserved} plan that explains the given inputs to the learning task. The intuition behind the compilation is that a solution to the resulting classical planning task is a sequence of actions that:

\begin{enumerate}
\item {\bf Programs the action model $\mathcal{M}'$}. A solution plan starts with a {\em prefix} that, for each $\xi\in\mathcal{M}$, determines which fluents $f\in F_v(\xi)$ belong to its $pre(\xi)$, $del(\xi)$ and $add(\xi)$ sets.
\item {\bf Validates the action model $\mathcal{M}'$}. The solution plan continues with a postfix that reproduces the given input knowledge (the available observations of the plan executions) with the programmed action model $\mathcal{M}'$.
\end{enumerate}

@@

\subsection{Learning from from observations of plan executions}
Here we formalize the compilation for learning \strips\ action models with classical planning. Given a learning task $\Lambda=\tup{\mathcal{M},\Psi,\mathcal{T}}$ the compilation outputs a classical planning task $P_{\Lambda}=\tup{F_{\Lambda},A_{\Lambda},I_{\Lambda},G_{\Lambda}}$:
\begin{itemize}
\item $F_{\Lambda}$ contains:
\begin{itemize}
\item The set of fluents $F$ built instantiating the predicates $\Psi$ with the objects $\Omega$ that appear in the input observations, i.e. the blocks {\tt\small A} and {\tt\small B} in the example of Figure~\ref{fig:example-plans}. Formally, $\Omega=\bigcup_{\small s\in\mathcal{O}} obj(s)$, where $obj$ is a function that returns the objects that appear in a given state.
\item Fluents $pre_f(\xi)$, $del_f(\xi)$ and $add_f(\xi)$, for every $f\in F_v(\xi)$, that represent the programmed action model. If a fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ holds, it means that $f$ is a precondition/negative/positive effect in the schema $\xi\in \mathcal{M}'$. For instance, the preconditions of the $stack$ schema (Figure~\ref{fig:stack}) are represented by the pair of fluents {\small\tt pre\_holding\_stack\_$v_1$} and {\small\tt pre\_clear\_stack\_$v_2$} set to True.
\item Fluents $F_{\pi}=\{plan(name(a_j),\Omega^{ar(a_j)},j)\}_{\small 1\leq j\leq |\pi|}$ to code the $j^{th}$ step of the observed plan $\pi=\tup{a_1, \ldots, a_n}$ that corresponds to action $a_j$. The static facts $next_{j,j+1}$ and the fluents $at_j$, {\small $1\leq j< |\pi|$}, are also added to iterate through the steps of plan $\pi$
  \item The fluents $mode_{prog}$ and $mode_{val}$ to indicate whether the operator schemas are programmed or validated, and the fluents $\{test_i\}_{1\leq i\leq n}$, indicating the state observation $s_i\in\mathcal{T}$ where the action model is validated.
\end{itemize}
\item $I_{\Lambda}$ encodes the first state observation, $s_0\subseteq F$ and sets to true $mode_{prog}$ as well as the fluents $F_{\pi}$ plus fluents $at_1$ and $\{next_{j,j+1}\}$, {\small $1\leq j<|\pi|$}, for tracking the plan step where the action model is validated. Our compilation assumes that initially, operator schemas are programmed with every possible precondition (the most specific learning hypothesis), no negative effect and no positive effect. Therefore fluents $pre_f(\xi)$, for every $f\in F_v(\xi)$, hold also at the initial state.

@@

\begin{figure}[hbt!]
\begin{scriptsize}
\begin{verbatim}
(:action apply_stack
  :parameters (?o1 - object ?o2 - object)
  :precondition
   (and (or (not (pre_on_stack_v1_v1)) (on ?o1 ?o1))
        (or (not (pre_on_stack_v1_v2)) (on ?o1 ?o2))
        (or (not (pre_on_stack_v2_v1)) (on ?o2 ?o1))
        (or (not (pre_on_stack_v2_v2)) (on ?o2 ?o2))
        (or (not (pre_ontable_stack_v1)) (ontable ?o1))
        (or (not (pre_ontable_stack_v2)) (ontable ?o2))
        (or (not (pre_clear_stack_v1)) (clear ?o1))
        (or (not (pre_clear_stack_v2)) (clear ?o2))
        (or (not (pre_holding_stack_v1)) (holding ?o1))
        (or (not (pre_holding_stack_v2)) (holding ?o2))
        (or (not (pre_handempty_stack)) (handempty)))
  :effect
   (and (when (del_on_stack_v1_v1) (not (on ?o1 ?o1)))
        (when (del_on_stack_v1_v2) (not (on ?o1 ?o2)))
        (when (del_on_stack_v2_v1) (not (on ?o2 ?o1)))
        (when (del_on_stack_v2_v2) (not (on ?o2 ?o2)))
        (when (del_ontable_stack_v1) (not (ontable ?o1)))
        (when (del_ontable_stack_v2) (not (ontable ?o2)))
        (when (del_clear_stack_v1) (not (clear ?o1)))
        (when (del_clear_stack_v2) (not (clear ?o2)))
        (when (del_holding_stack_v1) (not (holding ?o1)))
        (when (del_holding_stack_v2) (not (holding ?o2)))
        (when (del_handempty_stack) (not (handempty)))
        (when (add_on_stack_v1_v1) (on ?o1 ?o1))
        (when (add_on_stack_v1_v2) (on ?o1 ?o2))
        (when (add_on_stack_v2_v1) (on ?o2 ?o1))
        (when (add_on_stack_v2_v2) (on ?o2 ?o2))
        (when (add_ontable_stack_v1) (ontable ?o1))
        (when (add_ontable_stack_v2) (ontable ?o2))
        (when (add_clear_stack_v1) (clear ?o1))
        (when (add_clear_stack_v2) (clear ?o2))
        (when (add_holding_stack_v1) (holding ?o1))
        (when (add_holding_stack_v2) (holding ?o2))
        (when (add_handempty_stack) (handempty))
        (when (modeProg) (not (modeProg)))))
\end{verbatim}
\end{scriptsize}
 \caption{\small PDDL action for applying an already programmed schema $stack$ (implications are coded as disjunctions).}
\label{fig:compilation}
\end{figure}
The actions for {\em applying} an already programmed operator includes the following extra conditional effects $\{at_{j},plan(name(a_j),\Omega^{ar(a_j)},j)\}\rhd\{\neg at_{j},at_{j+1}\}_{\forall j\in [1,n]}$ for advancing to the next plan step. This mechanism ensures that $\mathsf{apply_{\xi,\omega}}$ actions are applied, exclusively, in the same order as in $\mathcal{T}$.

\item Actions for {\em validating} the partially observed state $s_i\in\mathcal{T}$, {\tt\small $1\leq i< n$}.
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{validate_{i}})=&s_i\cup\{test_j\}_{j\in 1\leq j<i}\cup\{\neg test_j\}_{j\in i\leq j\leq |\mathcal{O}|}\cup \{mode_{val}\},\\
\cond(\mathsf{validate_{i}})=&\{\emptyset\}\rhd\{test_i,\neg mode_{val}\}.
\end{align*}
\end{small}
\end{enumerate}
\end{itemize}

Further, known preconditions and effects (that is, a partially specified \strips\ action model) can be encoded as fluents $pre_f(\xi)$, $del_f(\xi)$ and $add_f(\xi)$ set to true at the initial state $I_{\Lambda}$. In this case, the corresponding programming actions, $\mathsf{programPre_{f,\xi}}$ and $\mathsf{programEff_{f,\xi}}$, become unnecessary and are removed from $A_{\Lambda}$ making the classical planning task $P_{\Lambda}$ easier to be solved. When a {\em fully} or {\em partially specified} \strips\ action model $\mathcal{M}$ is given, the compilation validates whether the observation of the plan execution follows the given model:
\begin{itemize}
\item $\mathcal{M}$ is proved to be a {\em valid} \strips\ action model for the given input data if a solution plan for $P_{\Lambda}$ can be found.
\item $\mathcal{M}$ is proved to be a {\em invalid} \strips\ action model for the given input data if $P_{\Lambda}$ is unsolvable. This means that $\mathcal{M}$ cannot be compliant with the given observation of the plan execution.
\end{itemize}
This feature of our compilation is beyond the functionality of VAL, the plan validation tool~\cite{howey2004val}, because VAL requires (1) a full plan and (2), a full action model for plan validation.

The classical plan of Figure~\ref{fig:plan-lplan} shows a solution to a learning task $\Lambda=\tup{\mathcal{M},\Psi,\mathcal{T}}$ for getting the {\em blocksworld} action model where operator schemes for {\tt\small pickup}, {\tt\small putdown} and {\tt\small unstack} are specified in $\mathcal{M}$. This plan programs and validates the operator schema {\tt\small stack} from {\em blocksworld}, using the plan $\pi$ and the two state observations $s_0$ and $s_4$ shown in Figure~\ref{fig:example-plans}. Plan steps $[0,8]$ program the preconditions of the {\tt\small stack} operator, steps $[9,13]$ program the operator effects and steps $[14,18]$ validate the programmed operators following the four-action plan shown in Figure~\ref{fig:example-plans}.

\begin{figure}[hbt!]
{\footnotesize\tt
     {\bf 00} : (program\_pre\_clear\_stack\_v1)\\
     01 : (program\_pre\_handempty\_stack)\\
     02 : (program\_pre\_holding\_stack\_v2)\\
     03 : (program\_pre\_on\_stack\_v1\_v1)\\
     04 : (program\_pre\_on\_stack\_v1\_v2)\\
     05 : (program\_pre\_on\_stack\_v2\_v1)\\
     06 : (program\_pre\_on\_stack\_v2\_v2)\\
     07 : (program\_pre\_ontable\_stack\_v1)\\
     08 : (program\_pre\_ontable\_stack\_v2)\\
     {\bf 09} : (program\_eff\_clear\_stack\_v1)\\
    10 : (program\_eff\_clear\_stack\_v2)\\
    11 : (program\_eff\_handempty\_stack)\\
    12 : (program\_eff\_holding\_stack\_v1)\\
    13 : (program\_eff\_on\_stack\_v1\_v2)\\
    {\bf 14} : (apply\_unstack blockB blockA i1 i2)\\
    15 : (apply\_putdown blockB i2 i3)\\
    16 : (apply\_pickup blockA i3 i4)\\
    17 : (apply\_stack blockA blockB i4 i5)\\
    {\bf 18} : (validate\_1)
}
 \caption{\small Plan for programming and validating the $stack$ schema using plan $\pi$ and state observations $\mathcal{O}$ (shown in Figure~\ref{fig:example-plans}) as well as previously specified operator schemes for $pickup$, $putdown$ and $unstack$.}
\label{fig:plan-lplan}
\end{figure}

Note that the $P_{\Lambda}$ compilation is flexible to an unbound number of {\em missing states} or {\em missing actions} in the input plan trace $\mathcal{T}$. If any reference to the $mode_{val}$ fluent is removed the compilation can learn \strips\ action models when there are missing state observations in $\mathcal{T}$. Likewise some of the actions in $\mathcal{T}$ may be unknown so they are not specified in $I_{\Lambda}$. In both cases the output planning becomes harder since the plan horizon is no longer bound (classical planning is PSPACE complete while SAT is NP) and the classical planner must determine how many {\em apply} actions are necessary between any two state observations, i.e. between the application of two {\em validate} actions.

Last but not least we explain how to address learning \strips\ action models from the observation of the execution of multiple plans $\Pi=\{\pi_1,\ldots,\pi_{\tau}\}$, {\tt\small $1\leq t\leq \tau$}. Let us first define a set of classical planning instances $P_t=\tup{F,\emptyset,I_t,G_t}$ that belong to the same planning frame (i.e. same fluents and actions but different initial states and goals). The set of actions, $A=\emptyset$, is empty because the action model is initially unknown. Finally, the initial state $I_t$ is given by the state $s_0^t$ and the plan $\pi_t$, and the goals $G_t$ are defined by the state $s_n^t$. Addressing the learning task $\Lambda=\tup{\mathcal{M},\Psi,\mathcal{T}}$ where $\mathcal{T}=\{t_1,\ldots,t_{\tau}\}$ requires introducing a small modification to our compilation. In particular, the actions in $P_{\Lambda}$ for {\em validating} the plan $\pi_t\in\Pi$, {\tt\small $1\leq t\leq \tau$} resets the current state, and the current plan, and are now defined as:
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{validate_{t}})=&G_t\cup\{test_j\}_{1\leq j<t}\cup\{\neg test_j\}_{t\leq j\leq \tau}\cup \{\neg mode_{prog}\},\\
\cond(\mathsf{validate_{t}})=&\{\emptyset\}\rhd\{test_t\} \cup \{\neg f\}_{\forall f\in G_t, f \notin I_{t+1}}\cup \{f\}_{\forall f\in I_{t+1}, f \notin G_t}.
\end{align*}
\end{small}

@@

\begin{proof}[Proof sketch]
\begin{small}
Once operator schemas $\mathcal{M}'$ are programmed, they can only be applied and validated, because of the $mode_{prog}$ fluent. In addition, $P_{\Lambda}$ is only solvable if fluents $\{test_i\}$, {\small $1\leq i\leq n$} hold at the last reached state. These goals can only be achieved executing an applicable sequence of programmed operator schemas that reaches every state $s_i\in\mathcal{T}$, starting from the corresponding initial state and following the sequence of actions defined by the plans in $\Pi$. This means that the programmed action model $\mathcal{M}'$ complies with the provided input knowledge and hence, solves $\Lambda$.
\end{small}
\end{proof}

@@

The size of the classical planning task $P_{\Lambda}$ output by the compilation depends on:
\begin{itemize}
\item The arity of the actions headers in $\mathcal{M}$ and the predicates $\Psi$ that are given as input to the $\Lambda$ learning task. The larger these numbers, the larger the size of the $F_v(\xi)$ sets. This is the term that dominates the compilation size because it defines the $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ fluents set and the corresponding set of {\em programming} actions.
\item The number of given state observations. The larger $n$, the more $test_i$ fluents and $\mathsf{validate_{i}}$ actions in $P_{\Lambda}$.
\end{itemize}

\subsection{Optimizing the compilation with background knowledge}
A distinctive feature of Inductive Logic Programming (ILP) is that ILP can leverage {\em background knowledge} to learn logic programs from data~\cite{muggleton1994inductive}. Inspired by ILP, we show that our approach for the learning of \strips\ action models can also leverage {\em background knowledge} in this case to optimize the performance of the $P_{\Lambda}$ compilation.

@@

{\em State invariants} is a kind of state-constraints useful for computing more compact state representations~\cite{helmert2009concise} or making {\em satisfiability planning} and {\em backward search} more efficient~\cite{rintanen2014madagascar,alcazar2015reminder}. Given a classical planning problem $P=\tup{F,A,I,G}$, a {\em state invariant} is a formula $\phi$ that holds at the initial state of a given classical planning problem, $I\models \phi$, and at every state $s$, built from $F$, that is reachable from $I$.

@@

An interesting contribution of our compilation is that the validation of an action model can also be done with {\em state constraints}. Given $\Phi$, a set of either state or trajectory constraints, our validate actions can be adapted to check that the learned model satisfy every constraint $\phi\in\Phi$:
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{validate_{i}})=&\phi\cup\{test_j\}_{j\in 1\leq j<i}\cup\{\neg test_j\}_{j\in i\leq j\leq |\Phi|}\cup \{mode_{val}\},\\
\cond(\mathsf{validate_{i}})=&\{\emptyset\}\rhd\{test_i,\neg mode_{val}\}.
\end{align*}
\end{small}

This redefinition of validate actions applies also to trajectory constraints because LTL formulae can be represented using classical action preconditions and goals by encoding the Non-deterministic B\"{u}chi Automaton (NBA), that is equivalent to the corresponding LTL formula, as part of the classical planning tasks~\cite{baier2006planning}.

>>>>

Our approach for addressing a $\Lambda$ learning task is compiling it into a classical planning task $P_{\Lambda}$ with conditional effects. The intuition behind the compilation is that a solution to the resulting classical planning task is a sequence of actions that:

\begin{enumerate}
\item {\bf Programs the action model $\mathcal{M}'$}. A solution plan starts with a {\em prefix} that, for each $\xi\in\mathcal{M}$, determines which fluents $f\in F_v(\xi)$ belong to its $pre(\xi)$, $del(\xi)$ and $add(\xi)$ sets.
\item {\bf Validates the action model $\mathcal{M}'$}. The solution plan continues with a {\em postfix} that reproduces the given input knowledge (the available observations of the plan executions) with the programmed action model $\mathcal{M}'$.
\end{enumerate}

@@

\subsection{Learning from from observations of plan executions}
Here we formalize the compilation for learning \strips\ action models with classical planning. Given a learning task $\Lambda=\tup{\mathcal{M},\Psi,\mathcal{T}}$ the compilation outputs a classical planning task $P_{\Lambda}=\tup{F_{\Lambda},A_{\Lambda},I_{\Lambda},G_{\Lambda}}$:
\begin{itemize}
\item $F_{\Lambda}$ contains:
\begin{itemize}
\item The set of fluents $F$ built instantiating the predicates $\Psi$ with the objects $\Omega$ that appear in the plan trace given as input, i.e. the blocks {\tt\small A} and {\tt\small B} in the example of Figure~\ref{fig:example-plans}. Formally, $\Omega=\bigcup_{\small s\in\mathcal{O}} obj(s)$, where $obj$ is a function that returns the objects that appear in a given state.
\item Fluents $pre_f(\xi)$, $del_f(\xi)$ and $add_f(\xi)$, for every $f\in F_v(\xi)$, that represent the programmed action model. If a fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ holds, it means that $f$ is a precondition/negative/positive effect in the schema $\xi\in \mathcal{M}'$. For instance, the preconditions of the $stack$ schema (Figure~\ref{fig:stack}) are represented by the pair of fluents {\small\tt pre\_holding\_stack\_$v_1$} and {\small\tt pre\_clear\_stack\_$v_2$} set to True.
\item The fluents $F_{\pi}=\{plan(name(a_i),\Omega^{ar(a_i)},i)\}_{\small 1\leq i\leq n}$ to code the $i^{th}$ step of the observed plan $\pi=\tup{a_1, \ldots, a_n}$ that corresponds to action $a_i$. The static facts $next_{i,i+1}$ and the fluents $at_i$, {\small $1\leq i< n$}, are also added to iterate through the $n$ steps of plan $\pi$
  \item The fluents $mode_{prog}$ and $mode_{val}$ to indicate whether the operator schemas are programmed or validated, and the fluents $\{test_i\}_{1\leq i\leq n}$, indicating the state observation $s_i\in\mathcal{T}$ where the action model is validated.
\end{itemize}
\item $I_{\Lambda}$ encodes the first state observation, $s_0\subseteq F$ and sets to true $mode_{prog}$ as well as the fluents $F_{\pi}$ plus fluents $at_1$ and $\{next_{i,i+1}\}$, {\small $1\leq i<n$}, for tracking the plan step where the action model is validated. Our compilation assumes that initially, operator schemas are programmed with every possible precondition (the most specific learning hypothesis), no negative effect and no positive effect. Therefore fluents $pre_f(\xi)$, for every $f\in F_v(\xi)$, hold also at the initial state.

@@

\begin{figure}[hbt!]
\begin{scriptsize}
\begin{verbatim}
(:action apply_stack
  :parameters (?o1 - object ?o2 - object)
  :precondition
   (and (or (not (pre_on_stack_v1_v1)) (on ?o1 ?o1))
        (or (not (pre_on_stack_v1_v2)) (on ?o1 ?o2))
        (or (not (pre_on_stack_v2_v1)) (on ?o2 ?o1))
        (or (not (pre_on_stack_v2_v2)) (on ?o2 ?o2))
        (or (not (pre_ontable_stack_v1)) (ontable ?o1))
        (or (not (pre_ontable_stack_v2)) (ontable ?o2))
        (or (not (pre_clear_stack_v1)) (clear ?o1))
        (or (not (pre_clear_stack_v2)) (clear ?o2))
        (or (not (pre_holding_stack_v1)) (holding ?o1))
        (or (not (pre_holding_stack_v2)) (holding ?o2))
        (or (not (pre_handempty_stack)) (handempty)))
  :effect
   (and (when (del_on_stack_v1_v1) (not (on ?o1 ?o1)))
        (when (del_on_stack_v1_v2) (not (on ?o1 ?o2)))
        (when (del_on_stack_v2_v1) (not (on ?o2 ?o1)))
        (when (del_on_stack_v2_v2) (not (on ?o2 ?o2)))
        (when (del_ontable_stack_v1) (not (ontable ?o1)))
        (when (del_ontable_stack_v2) (not (ontable ?o2)))
        (when (del_clear_stack_v1) (not (clear ?o1)))
        (when (del_clear_stack_v2) (not (clear ?o2)))
        (when (del_holding_stack_v1) (not (holding ?o1)))
        (when (del_holding_stack_v2) (not (holding ?o2)))
        (when (del_handempty_stack) (not (handempty)))
        (when (add_on_stack_v1_v1) (on ?o1 ?o1))
        (when (add_on_stack_v1_v2) (on ?o1 ?o2))
        (when (add_on_stack_v2_v1) (on ?o2 ?o1))
        (when (add_on_stack_v2_v2) (on ?o2 ?o2))
        (when (add_ontable_stack_v1) (ontable ?o1))
        (when (add_ontable_stack_v2) (ontable ?o2))
        (when (add_clear_stack_v1) (clear ?o1))
        (when (add_clear_stack_v2) (clear ?o2))
        (when (add_holding_stack_v1) (holding ?o1))
        (when (add_holding_stack_v2) (holding ?o2))
        (when (add_handempty_stack) (handempty))
        (when (modeProg) (not (modeProg)))))
\end{verbatim}
\end{scriptsize}
 \caption{\small PDDL action for applying an already programmed schema $stack$ (implications are coded as disjunctions).}
\label{fig:compilation}
\end{figure}
These $\mathsf{apply_{\xi,\omega}}$ actions includes the extra conditional effects $\{at_{i},plan(name(a_i),\Omega^{ar(a_i)},i)\}\rhd\{\neg at_{i},at_{i+1}\}_{\forall i\in [1,n]}$ to validate that actions are applied, exclusively, in the same order as in $\mathcal{T}$.

\item Actions for {\em validating} the partially observed state $s_i\in\mathcal{T}$, {\tt\small $1\leq i< n$}.
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{validate_{i}})=&s_i\cup\{test_j\}_{j\in 1\leq j<i}\cup\{\neg test_j\}_{j\in i\leq j\leq n}\cup \{mode_{val}\},\\
\cond(\mathsf{validate_{i}})=&\{\emptyset\}\rhd\{test_i,\neg mode_{val}\}.
\end{align*}
\end{small}
\end{enumerate}
\end{itemize}

Known preconditions and effects (that is, a partially specified \strips\ action model) can be encoded as fluents $pre_f(\xi)$, $del_f(\xi)$ and $add_f(\xi)$ set to true at the initial state $I_{\Lambda}$. In this case, the corresponding programming actions, $\mathsf{programPre_{f,\xi}}$ and $\mathsf{programEff_{f,\xi}}$, become unnecessary and are removed from $A_{\Lambda}$ making the classical planning task $P_{\Lambda}$ easier to be solved. When a {\em fully} or {\em partially specified} \strips\ action model $\mathcal{M}$ is given, the $P_{\Lambda}$ compilation validates whether the observation of the plan execution follows the given model:
\begin{itemize}
\item $\mathcal{M}$ is proved to be a {\em valid} \strips\ action model for the given input data in $\mathcal{T}$ iff a solution plan for $P_{\Lambda}$ can be found.
\item $\mathcal{M}$ is proved to be a {\em invalid} \strips\ action model for the given input data $\mathcal{T}$ iff $P_{\Lambda}$ is unsolvable. This means that $\mathcal{M}$ cannot be compliant with the given observation of the plan execution.
\end{itemize}
This feature of our compilation is beyond the functionality of VAL, the plan validation tool~\cite{howey2004val}, because VAL requires (1) a full plan and (2), a full action model for plan validation. On the other hand our $P_{\Lambda}$ compilation can test {\em model validation} with a partial (or even an empty) action model and a partially observed plan trace.

The classical plan of Figure~\ref{fig:plan-lplan} shows a solution to the classical planning task $P_{\Lambda}$ that encodes a $\Lambda=\tup{\mathcal{M},\Psi,\mathcal{T}}$ learning task for getting the {\em blocksworld} action model where operator schemes for {\tt\small pickup}, {\tt\small putdown} and {\tt\small unstack} are specified in $\mathcal{M}$. This plan programs and validates the operator schema {\tt\small stack} from {\em blocksworld}, using the plan $\pi$ and the two state observations $s_0$ and $s_4$ shown in Figure~\ref{fig:example-plans}. Plan steps $[0,8]$ program the preconditions of the {\tt\small stack} operator, steps $[9,13]$ program the operator effects and steps $[14,18]$ validate the programmed operators following the four-action plan shown in Figure~\ref{fig:example-plans}.

\begin{figure}[hbt!]
{\footnotesize\tt
     {\bf 00} : (program\_pre\_clear\_stack\_v1)\\
     01 : (program\_pre\_handempty\_stack)\\
     02 : (program\_pre\_holding\_stack\_v2)\\
     03 : (program\_pre\_on\_stack\_v1\_v1)\\
     04 : (program\_pre\_on\_stack\_v1\_v2)\\
     05 : (program\_pre\_on\_stack\_v2\_v1)\\
     06 : (program\_pre\_on\_stack\_v2\_v2)\\
     07 : (program\_pre\_ontable\_stack\_v1)\\
     08 : (program\_pre\_ontable\_stack\_v2)\\
     {\bf 09} : (program\_eff\_clear\_stack\_v1)\\
    10 : (program\_eff\_clear\_stack\_v2)\\
    11 : (program\_eff\_handempty\_stack)\\
    12 : (program\_eff\_holding\_stack\_v1)\\
    13 : (program\_eff\_on\_stack\_v1\_v2)\\
    {\bf 14} : (apply\_unstack blockB blockA i1 i2)\\
    15 : (apply\_putdown blockB i2 i3)\\
    16 : (apply\_pickup blockA i3 i4)\\
    17 : (apply\_stack blockA blockB i4 i5)\\
    {\bf 18} : (validate\_1)
}
 \caption{\small Plan for programming and validating the $stack$ schema (using plan $\pi$ and state observations shown in Figure~\ref{fig:example-plans}) as well as previously specified operator schemes for $pickup$, $putdown$ and $unstack$.}
\label{fig:plan-lplan}
\end{figure}

Note that the $P_{\Lambda}$ compilation is flexible to an unbound number of {\em missing states} or {\em missing actions} in the input plan trace $\mathcal{T}$. If any reference to the $mode_{val}$ fluent is removed the compilation can learn \strips\ action models when there are missing state observations in $\mathcal{T}$. Likewise some of the actions in $\mathcal{T}$ may be unknown so they are not specified in $I_{\Lambda}$. In both cases the output planning problem becomes harder since the plan horizon is no longer bound and the classical planner must determine how many {\em apply} actions are necessary between any two state observations, i.e. between the application of two {\em validate} actions (classical planning is {\em PSPACE complete} while classical planning with a fixed plan horizon is {\em NP}~\cite{rintanen:SATplannning:AI2012}).

Last but not least we explain how to address learning \strips\ action models from the observation of the execution of multiple plans $\Pi=\{\pi_1,\ldots,\pi_{\tau}\}$, {\tt\small $1\leq t\leq \tau$}. Let us first define a set of classical planning instances $P_t=\tup{F,\emptyset,I_t,G_t}$ that belong to the same planning frame (i.e. same fluents and actions but different initial states and goals). The set of actions, $A=\emptyset$, is empty because the action model is initially unknown. Finally, the initial state $I_t$ is given by the state $s_0^t$ and the plan $\pi_t$, and the goals $G_t$ are defined by the state $s_n^t$. Addressing the learning task $\Lambda=\tup{\mathcal{M},\Psi,\mathcal{T}}$ where $\mathcal{T}=\{t_1,\ldots,t_{\tau}\}$ requires introducing a small modification to our compilation. In particular, the actions in $P_{\Lambda}$ for {\em validating} the plan $\pi_t\in\Pi$, {\tt\small $1\leq t\leq \tau$} resets the current state, and the current plan, and are now redefined as:
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{validate_{t}})=&G_t\cup\{test_j\}_{1\leq j<t}\cup\{\neg test_j\}_{t\leq j\leq \tau}\cup \{\neg mode_{prog}\},\\
\cond(\mathsf{validate_{t}})=&\{\emptyset\}\rhd\{test_t\} \cup \{\neg f\}_{\forall f\in G_t, f \notin I_{t+1}}\cup \{f\}_{\forall f\in I_{t+1}, f \notin G_t}.
\end{align*}
\end{small}

@@

\begin{proof}[Proof sketch]
\begin{small}
Once operator schemas $\mathcal{M}'$ are programmed, they can only be applied and validated, because of the $mode_{prog}$ fluent. In addition, $P_{\Lambda}$ is only solvable if fluents $\{test_i\}$, {\small $1\leq i\leq n$} hold at the last reached state. These goals can only be achieved executing an applicable sequence of programmed operator schemas that reaches every state $s_i\in\mathcal{T}$, starting from the corresponding initial state and following the sequence of actions defined by $\mathcal{T}$. This means that the programmed action model $\mathcal{M}'$ complies with the provided input knowledge and hence, solves $\Lambda$.
\end{small}
\end{proof}

@@

The size of the classical planning task $P_{\Lambda}$ output by the compilation depends on:
\begin{itemize}
\item The arity of the actions headers in $\mathcal{M}$ and the predicates $\Psi$ that are given as input to the $\Lambda$ learning task. The larger these numbers, the larger the size of the $F_v(\xi)$ sets. This is the term that dominates the compilation size because it defines the $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ fluents set and the corresponding set of {\em programming} actions.
\item The number of given observations in $\mathcal{T}$. The larger $n$, the more $test_i$ fluents and $\mathsf{validate_{i}}$ actions in $P_{\Lambda}$.
\end{itemize}

\subsection{Tuning the compilation with background knowledge}
A distinctive feature of Inductive Logic Programming (ILP) is that ILP can leverage {\em background knowledge} to learn logic programs from data~\cite{muggleton1994inductive}. Inspired by ILP, we show that our approach for the learning of \strips\ action models can also leverage {\em background knowledge} in this case to optimize the performance of the $P_{\Lambda}$ compilation.

@@

{\em State invariants} is a kind of state-constraints useful for computing more compact state representations~\cite{helmert2009concise} or making {\em satisfiability planning} and {\em backward search} more efficient~\cite{rintanen2014madagascar,alcazar2015reminder}. Given a classical planning problem $P=\tup{F,A,I,G}$, a {\em state invariant} is a formula $\phi$ that holds at the initial state of a given classical planning problem, $I\models \phi$, and at every state $s$, built from $F$, that is reachable from $I$ by applying actions in $A$.

@@

An interesting contribution of our compilation is that the validation of an action model can also be done with {\em state constraints}. Given $\Phi$, a set of either state or trajectory constraints, our validate actions can be adapted to check also that the learned model satisfy a constraint $\phi\in\Phi$:
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{validate_{i}})=&\phi\cup\{test_j\}_{j\in 1\leq j<i}\cup\{\neg test_j\}_{j\in i\leq j\leq |\Phi|}\cup \{mode_{val}\},\\
\cond(\mathsf{validate_{i}})=&\{\emptyset\}\rhd\{test_i,\neg mode_{val}\}.
\end{align*}
\end{small}

This redefinition of validate actions applies also to {\em trajectory constraints} because LTL formulae can be represented using classical action preconditions and goals by encoding the Non-deterministic B\"{u}chi Automaton (NBA), that is equivalent to the corresponding LTL formula, as part of the classical planning tasks~\cite{baier2006planning}.
